---
title: "3. Write a Function!"
description: |
  Demonstration of how to write a function with personality research data.
author:
  - name: Raleigh Goodwin, Vinita Vader
date: 05-28-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## Introduction

# Ipsatization

This is a tutorial on using functional programming to solve specific problems in research. This tutorial addresses the issue of Ipsatization which consists of methods of data transformation used in Personality Psychology and Social Psychology research. Ipsatization transforms each participant's ratings relative to their average response such that the total and the average of the participant's scores across all items in the data set are zero (or another constant for all people) ([Greer and Dunlap, 1997](https://psycnet.apa.org/record/1997-06270-006)).

Packages such as [`multicon`](https://cran.r-project.org/web/packages/multicon/multicon.pdf) have built functions like [`ipsatize()`](https://rdrr.io/cran/multicon/man/ipsatize.html) which enable standardizing rows of the dataframes being studied. However it does not address the various types of ipsative scorings available for carrying out different transformations.

An important aspect of using data transformations involves understanding the relationship between raw data and transformed data. The purpose of the function built here will be to address this specific issue.

# Loading Libraries

Before we get started, we need to load the libraries necessary to complete this tutorial. Loading the entire library may not be always necessary, especially if you intend to use it only once. This will be the case for `rio`, `here`, and `knitr` in this tutorial, so you may choose not to load them here if you'd like.

```{r}
library(tidyverse)
library(purrr)
library(rio) # optional
library(here) #optional
library(knitr) #optional
```

# About the Data

For this tutorial, we will be working with a dataset containing the Ten Item Personality Inventory (TIPI; [Gosling, S. D., Rentfrow, P. J., & Swann, W. B., Jr., 2003](https://psycnet.apa.org/record/2003-09807-003)), a brief measure of the Big Five Personality Domains ([Goldberg, 1993](https://psycnet.apa.org/record/1993-17546-001)). Each item asks respondents to rate themselves on attributes (e.g., extroverted, critical, anxious, calm, etc.) using a Likert scale ranging from 1 to 7, wherein:\
\
1 = "Disagree strongly"\
2 = "Disagree moderately"\
3 = "Disagree a little"\
4 = "Neither agree nor disagree"\
5 = "Agree a little"\
6 = "Agree moderately"\
7 = "Agree strongly"

This particular dataset contains observations from N=2495 individuals who completed, among many other measures, the 10 TIPI items in 2016. Other variables included the Generic Conspiracist Beliefs Scale ([Brotherton et al., 2013](https://www.frontiersin.org/articles/10.3389/fpsyg.2013.00279/full)), various response time metrics, a vocabulary validity check, and demographics.

### Importing the Data

When importing data, two important things to keep in mind are your working directory and reproducibility. Where you save your files can impact the ease at which you can call them; you'll have the best luck saving data files of interest within the corresponding R Project. `rio`'s `import()` function provides an easy method for importing data files, including the ability to set the class of the data to tibble, which helps to retain the data in a format which is more amenable to data manipulation in `tidyverse`. To enhance reproducibility across different devices, we'll use the `here()` function within the [`here`](https://here.r-lib.org/) package when specifying our file path.

```{r}
# Import data
df_tipi <- rio::import(here::here("content/dataCT.csv"), setclass = "tibble")
```

For the current project, we'll only be working with the TIPI items, so to simplify the dataframe we're using, we can select only those columns.

```{r}
# Select desired variables
data <- df_tipi %>% 
  select(TIPI1:TIPI10)
```

Now we can take a look at the data we'll be working with. The `kable()` function from the `knitr` package helps to format the data into a neat table.

```{r}
# Take a look at the data
data %>% 
  head(n = 5) %>% 
  knitr::kable()
```

With the libraries loaded and data imported, we can now begin building our function.

## Building Functions

There are several ways in which one could go about building functions. The approach outlined here should be viewed as one of the several approaches to go about building functions.

As you think about building a function, keep in mind the purpose of why you set to build a function in the first place. Your function will ideally solve a problem specific to your analysis or can also be used by others to carry out their analyses.

Let's state the problem first: The difference between raw and ipsatized data has been studied to some extent leading to several debates amongst researchers questioning the utility of these methods. It is therefore important to look at correlations between the raw and ipsatized data.

Now that we understand the problem, let's think about how our function could address this problem. Here are a sequence of questions which will help you think about the function you intend to build.

__What is the goal of this function?__\
What do we need this function to do? For the current tutorial, we are writing a function that ipsatizes any dataset, meaning that it will compute the means of the rows and subtract the mean from every score in the respective rows. Ideally, it will produce output in the form of a list containing the raw and transformed (i.e., ipsatized) data.

```{r eval=FALSE}
ipsatize.MeanDev(dataframe) # Raleigh edit: I'm not sure what's going on in this chunk

>list
 $raw...
 $ipsatize...
```

__How can we achieve this goal for a specific dataset?__\
As we learned in previous posts, when taking a functional programming approach to this problem, we should first attempt to solve it within a specific case. Once we've done so, we can then consider generalizing to a function. For this tutorial, we will be solving the problem first with the TIPI dataset, and then we can apply that solution to build the final function.

__How can we break the function's goal into smaller tasks?__\
Most likely, we aren't just going to be writing one function in this tutorial. Ideally, a function should complete exactly one task; therefore, when we are attempting to build a function to complete a complicated task like ipsatization, we will need to write multiple simple functions and combine them. Thus, it can be helpful to first outline and think through each step of the process and eventually create a function for each step.

### Step 1: Creating an ID column

To ipsatize the data, we need to calculate each participant's mean response to the TIPI scale items and then subtract each response by that mean. This means we need to be able to conduct these operations by row rather than by column. One of the easier ways to do this is to use `pivot_longer()` to transform the data into a "longer" format. 

First, though, we need to create an ID for each participant that can then be used to identify their responses once the data is transformed.

```{r}
data_id <- data %>% 
  mutate(id = c(1:nrow(data))) # Create ID column

data_id %>% 
  head(n = 5) %>% # Take a look at the first 5 rows of the resulting dataframe
  knitr::kable() # Format the output table neatly
```

Now that we have an ID variable that can be used to identify each participant's responses, we can figure out how to create a column that calculates the mean of each rows using `pivot_longer()`.

```{r}
data_long <- data_id %>% 
  pivot_longer(cols = !id, names_to = "item", values_to = "response")

data_long %>% 
  head(n = 15) %>% 
  knitr::kable()
```

Instead of participants' responses being organized by row, all responses are now contained in one column and can be identified using the corresponding ID and Item values. We can use `tidyverse`'s `group_by()` function to group this dataframe by participant ID and then compute 1) the mean for each group and 2) the difference between each response and the mean of its group.

```{r}
data_long <- data_long %>% 
  group_by(id) %>% # Group by participant ID
  mutate(mean_row = mean(response, na.rm = TRUE), # Calculate participant mean
         deviation = response - mean_row) # Calculate individual response deviation from mean

data_long %>% 
  head(n = 15) %>% 
  knitr::kable()
```

Now, we can use `pivot_wider()` to transform the data back to its original format.

```{r}
data_wide <- data_long %>% 
  pivot_wider(names_from = item, values_from = deviation) %>% # something is going wrong here
  select(id, !mean_row) %>% 
   ungroup()

data_wide %>% 
  head(n = 15) %>% 
  knitr::kable()
```

That works! Now let's wrap this within a function. We will call it `add_id`

```{r}
add_id <- function(df) {
  df_id <- df %>% 
    mutate(id = row_number())
  df_id
}
```

Now let's check if it works.

```{r}
add_id(data)%>% 
  head(n = 5) %>% 
  knitr::kable()
```



__How do we build the functions specified in step 2?__\
We will focus on one function at a time. We are going to first build the ipsatize function. Let's break down the ipsatize function.\
i. What is the _name_ of the function? \
We will call it `ipsatize_MeanDev` as the type of ipsatization this function will perform is Mean deviation.\
ii. What will the the function _do_? \

Now we can wrap this into a function

```{r}
ipsatize_MD <- function(df){
  
 dfid <- df %>% 
   add_id()
  
 output <- dfid %>% 
  pivot_longer(cols = !id, names_to = "col_names") %>% 
  group_by(id) %>% 
  mutate(means_rowwise = mean(value, na.rm = T),
         value = value - means_rowwise) %>% 
  pivot_wider(names_from = col_names, values_from = value) %>% 
  select(id, !means_rowwise) %>% 
   ungroup()

  output
}

ipsatize_MD(data)
```

This works! We can add constraints to which type of variables can be added in the dataframe. For this we can build a function which only selects numeric columns.

```{r}
just_num <- function(df) {
  if(sum(purrr::map_lgl(df, is.numeric)) == 0) {
    stop("No numeric columns.")
  }
    else{
      df1 <- df[ , purrr::map_lgl(df, is.numeric)]
      df1
    }
}
```

In the above function, we specify that if the sum of all possible numeric columns in a dataframe is 0, we stop the loop and throw an error message else return the dataframe. This will ensure that all the columns in the dataframe we are working with are numeric. We are using `map_lgl` from the `purrr` package which maps the `is.numeric` function to every column in the dataframe.

Let's give this function a try. We know that the dataframe "data" has all columns of class numeric whereas `df_tipi` does not.

```{r}
ncol(just_num(df_tipi)) == (ncol(df_tipi))
ncol(just_num(data)) == (ncol(data))
```

This works!

Now we can add this to the main function.

```{r echo=TRUE}
ipsatize_MD <- function(df){
  
 df_clean <- df %>% 
   just_num() %>% 
   add_id()
  
 output <- df_clean %>% 
  pivot_longer(cols = !id, names_to = "col_names") %>% 
  group_by(id) %>% 
  mutate(means_rowwise = mean(value, na.rm = T),
         value = value - means_rowwise) %>% 
  pivot_wider(names_from = col_names, values_from = value) %>% 
  select(id, !means_rowwise) %>% 
   ungroup() %>% 
   select(-id)

  output
}

ipsatize_MD(data)
```

A closer look will tell you that we could benefit from renaming the column names of the ips dataframe within the list. The `paste` function can be helpful to do this.

```{r echo=TRUE}
rename_ips <- function(ipsatized_data){
  names(ipsatized_data) <- paste0(names(ipsatized_data), "_ips")
  return(ipsatized_data)
}

rename_ips(ipsatize_MD(data))
```

All we need to do now is wrap up the ipsatized data and raw data within a list.

```{r echo=TRUE}
ipsatize_MeanDev <- function(set){
  
  l <- list("ips" = rename_ips(ipsatize_MD(data)) ,                "raw" = data,
            "correlation.matrix" = cor(rename_ips(ipsatize_MD(data)), data))
  l
}

ips_raw_data <- ipsatize_MeanDev(data)
```

So now we have a list with the three dataframes. We can look at the correlations between raw and ipsatized data.

```{r echo=TRUE}
tibble(
  diag(ips_raw_data$correlation.matrix), colnames(ips_raw_data$ips), colnames(ips_raw_data$raw)
)
```

We can also plot raw and ipsatized data. For example, lets look at the item 1.

```{r}
TIPI_item1 <- data.frame(ips_raw_data$raw$TIPI1, ips_raw_data$ips$TIPI1_ips)

TIPI_item1 %>% 
  ggplot()+
   geom_density(aes(x=TIPI_item1[,1]), colour="red", fill = "red", alpha = .5) + 
   geom_density(aes(x=TIPI_item1[,2]), colour="blue", fill = "blue", alpha = .5) +
  theme_minimal()
```

# Things to work on 
1. Formatting - how to add links 
3. Plots - make them pretty, can add a vline for the mean, can change colors to a colorblind palette, can annotate 
4. conclusion